Simulated observations of young protoplanetary discs

### 2013-01-11 paper.tex
more figures a few tweaks to the text

\title[Young protoplanetary discs]{Simulated observations of young protoplanetary discs}
\author[Tom  Douglas, Paola Caselli, Et al.]{Tom Douglas$^{1}$\thanks{E-mail:
pytd@leeds.ac.uk}, Paola Caselli$^{1}$, Aaron Boley$^{2}$, Richard Durisen$^{3}$, Tom Hartquist$^{1}$, John Ilee$^{1}$, Jonathon Rawlings$^{4}$ \\
$^{1}$School of Physics and Astronomy, University of Leeds, Leeds LS2 9JT, UK \\
$^{2}$Department of Astronomy, University of Florida, 211 Bryant Space Center, PO Box 112055, USA\\
$^{3}$Department of Astronomy, Indiana University, 727 East 3rd Street, Swain West 319, Bloomington, IN 47405, USA\\
%$^{4}$Department of Physics & Astronomy, University College London, London WC1E 6BT, UK\\
}
\begin{document}

@@

The formation and early evolution of protoplanetary discs around solar-type and low-mass protostars has little observational support, despite of the fast growing list of theoretical models on the dynamical evolution of star forming dense cores (e.g. Krasnopolsky et al. 2011; Machida et al. 2011; Braiding \& Wardle 2012). The reason for this is that young protostars are surrounded by thick envelopes and power energetic outflows, thus observations of the young discs, predicted to have sizes of about 100\,AU and masses as large as 10\% the original core mass (e.g. Joos et al. 2012), are challenging. The use of sensitive interferometers is needed to achieve high angular resolution and spatially/spectrally disentangle the various (disc + envelope + outflow) components as well to filter out the extended emission tracing envelope material.

After the pioneer work of, e.g., Chandler et al. (1995), Brown et al. (2000), Looney et al. (2000), recent interferometric observations have discovered compact embedded discs in a sample of Class 0 (Andr\'e et al. 2000) sources, finding masses between 0.4 and  $>$1 M$_{\odot}$ (Jorgensen et al. 2007, 2009; Enoch et al. 2011). A 130\,AU disc was discovered toward a Class 0 source in Perseus, using Jansky Very Large Array (JVLA) observations of NH$_3$ (Choi et al. 2007). Pineda et al. (2012) observed methyl formate with the Atacama Millimeter and sub-millimeter Array (ALMA) and found evidence of rotation toward one of the proto-binary Class 0 sources embedded in IRAS 16293-2422. These observations are consistent with an almost edge-on disc. Persson et al. (2012) observed H$_2^{18}$O with ALMA toward the same source and found evidence of relatively large ($\ga$100\,K) excitation temperatures, as well as a HDO/H$_2$O abundance ratio close to that measured on Earth's oceans and Jupiter-family comets (Hartogh et al. 2011). When ALMA will be completed, it will be finally possible to spatially resolve these young discs and, for the first time, put stringent constraints on theoretical models. As full-operational ALMA is fast approaching, it is important to provide observational predictions based on dynamical models of young protoplanetary discs.

In this paper, we focus our attention to the young self-gravitating discs of Boley \& Durisen (2008), where episodic heating induced by spiral shocks is present, as this may be a good representation of the earliest phases of protoplanetary discs and an alternative to the young "static" discs studied by, e.g. Visser et al. (2009, 2011). As shown by Ilee et al. (2011), the spiral shocks cause desorption of volatiles from dust icy mantles and trigger gas-phase chemical reactions with activation energies, which are inactive at lower temperatures. These processes produce clear chemical signatures of the disc dynamics. With the use of 3D radiative transfer modeling, we perform simulated ALMA observations of the disc studied by Ilee et al. (2011) and identify the best tracers of the physical structure of self-gravitating discs. The physical, chemical and radiative transfer models are described in Sect\,\ref{sec:description_model}. Radiative transfer results are in Sect.\,\ref{sec:model_results}, while ALMA simulated observations are in Sec.\,\ref{sec:alma_predictions}. Discussions and conclusions can be found in Sect.\,\ref{sec:discussion}.

@@

The physical model used to simulate the emission from a young protoplanetary disc is a hybrid model obtained from embedding a gravitationally unstable protoplanetary disc  within an envelope with characteristics similar to the well-studied pre-stellar core L1544, as derived by Keto \& Caselli (2010; hereafter KC2010). The KC2010 model follows the dynamical, chemical and thermal evolution of a contracting Bonnor-Ebert (Bonnor 1956; Ebert 1957) sphere with total mass of 10\,M$_{\odot}$, until it reaches the density, temperature and velocity profiles which best match observations. The pre-stellar core model adopted here contains slight modifications due to the inclusion of oxygen cooling in the outer regions of the cloud, where CO is mostly photodissociated (Keto et al., in preparation), and it is shown in Fig.\,\ref{fig:l1544_model}.  [NOTE: BESIDES THE PROBLEM WITH THE FIGURE NUMBER, YOU NEED TO READJUST FIGURES, AS THIS ONE IS THE FIRST ONE THAT YOU DESCRIBE, SO IT SHOULD BE NUMBER 1] \newline

\begin{figure}
 \includegraphics[width=84mm]{Figures/model/L1544model_used_legend.eps}
 \label{fig:l1544_model}
 \caption{Model of the pre-stellar core L1544 used as the envelope of the young protoplanetary disc in the hybrid model. Showing temperature (red) and dust (green) temperature in kelvin, log number density (blue) in cm$^{-3}$ and inward velocity$\times$100 (cyan) in m$\,$s$^{-1}$. Adapted from Keto \& Caselli (2010) and Keto et al. (2012, in preparation).}
\end{figure}

\begin{figure*}
 \includegraphics[width=168mm]{Figures/model/rhoT2.eps}
 \label{rhoT} 
 \caption{Left: A 3D plot of log number density (cm$^{-3}$) showing the spiral structure in the xy plane and scale height of the disc. Right: The 3D temperature structure of the disc; regions cooler than 40 degrees are not shown in 3D, demonstrating the narrow central region containing hot material.}
\end{figure*}

@@

%-describe the chemistry  - refer to Ilee et al. which have been taken as input to the rad transfer (RT) code
Chemical abundances in the disc were taken from Ilee et. al (2011) which followed the changes of chemical abundances of trace particles moving through the disc as it evolved. The abundances of 125 species related by 1334 reactions were followed through the time evolution of the disc. These abundances were interpolated onto a 51$^3$ grid covering the disc with cells of size 2.2$\times$2.2$\times$0.22$\,$au.  From Ilee et al. (2011) we selected the three species which appear to trace different regions of the disc : the inner 30\,AU (OCS), the inner 60\,AU (H$_2$CO) and the region between $\simeq$20 and 60\,AU (HCO$^+$) and one which traces the entirity of the disc (C$^{18}$O). As explained by Ilee et al. (2011),  H$_2$CO and OCS mostly probe the central warm regions, where icy mantles evaporate, whereas HCO$^+$ preferentially traces the outer spiral pattern as in the central region it is destroyed by water molecules. The simple chemistry in the KC2010 model, adopted here as the envelope of the protoplanetary disc, does not provide detailed abundances of molecular species (besides CO and H$_2$O, see also Caselli et al. 2012). As discussed in the result section (Sec.\,\ref{sec:model_results}), rough guesses have been made based on values measured toward similar objects.

%NOTE: WE SHOULD TALK ABOUT THESE SPECIES AFTER WE SHOW WHICH SPECIES WE SELECT FOR THE RADIATIVE TRANSFER. 
%Abundances in the envelope model were as follows:\newline

%(tableify this?)
%HCO$^+$: As the H$_2$O profile from Caselli et al. (2012, submitted) scaled so that the maximum abundance is 10$^{-8}$\newline
%HNO: Constant abundance of 5$\times$10$^{-11}$\newline
%HCS$^+$: Constant abundance of 10$^{-11}$\newline
%OCS: Constant abundance of 10$^{-9}$\newline
%H$_2$CO: 1.5$\times$10$^{-9}$ decreased by a factor of 40 at radii less than 8250au (Young et al 2004)\newline
%CS:  3$\times$10$^{-9}$ decreased by a factor of 10,000 at radii less than 6700au (Tafalla, santiago-garcia and myers 2006)\newline
%CO: As the H2O profile from Caselli (2012, submitted) scaled so that the maximum abundance is CO maximum ***which is....***\newline

\subsection{The radiative transfer code} \label{subsec:radiative_transfer_code}
%-describe the RT used (LIME) 
The radiative transfer program used, LIME (Brinch \&Hogerheijde 2010), calculates line intensities based on a weighted sample of randomly chosen points in a continuous 3D model. The method of selecting these points is given in the grid-construction section. At each of these points, the density of the main collision partner (H$_2$), gas and dust temperatures, velocity, molecular abundances and turbulent velocity are taken from the model. These points are then smoothed by Lloyds algorithm (Lloyd 1982) in order to minimise the variation in distance between points whilst keeping the same underlying distribution. These points are then connected by Delaunay triangulation \footnote{In three dimensions this means that if four points are conneted into a tetrahedron, the sphere circumscribing these four points contains no other points. It can be shown that this connection is unique for a given set of points.} and it is down these paths that photon propagation is restricted (figures \ref{grid}). The level populations of the selected molecules are calculated at each of these points from collisional and radiative (de)excitation and the local radiation field is calculated. This is repeated 20 times with the populations of each level converging towards a single value. This number of iterations is sufficient for the signal to noise ratio of the level populations (as defined in Brinch \&Hogerheijde 2010) to exceed 1000 in 99\% of the points, ensuring that the simulation has converged on a stable level population. \newline
After 20 iterations of updateing the level populations the model is ray-traced in order to produce synthetic brighness maps. In order to minimise the ``blockyness'' of the output images, resulting from the grid construction, the average of ten seperate runs was taken (figure \ref{averages}).

\begin{figure}
 \includegraphics[width=84mm]{Figures/sim/average_comparison.eps}
 \label{averages}
 \caption{A plot showing the difference between a single LIME continuum image at 300$\,$GHz (right) and the average of ten such images from different runs of the same model in LIME (left).}
\end{figure}

@@

\begin{figure}
 \includegraphics[width=84mm]{Figures/model/lime_points_rz_histo.eps}
 \label{points}
 \caption{A 2D histogram of the point distribution throughout the model. The disc and envelope can be seen as two separate entities which have to be sampled using different point distributions}
\end{figure}

@@

Simulations are limited to molecules for which we have abundances in both the disc and envelope, and also have calculated Einstein and collisional co-efficients. The Molecular data used is from the lamda database (reference? http://home.strw.leidenuniv.nl/$\sim$moldata/). Lines for the molecules which satisfy these conditions were simulated with LIME for freqencies between 50$\sim$500$\,$GHz, with a focus on lines which could be observed in ALMA band 7 and molecules which are not typically seen in outflows. The results presented in this section represent those lines which show dectectable emission/absorption which can be used to trace either spiral structure or rotation.\newline

Transitions focused upon are CO 3-2 (and isotopologues) at 345.8GHz with an upper energy level of 33.2K, HCO+ 1-0 at 89.2GHZ and 4.3K, OCS 28-27 at 340.4GHz and 237.0K and H2CO $4_{04}$-3$_{03}$ at 290.6GHz and 34.9K {\bf maybe worth including a higher frequency line as well? maybe CS at ~650GHz}. Other molecules simulated but not shown include HCN, HNC, HNO, HCS$^+$ and CS.\newline

For the purpose of simulating observations the model was placed at roughly the distance of nearby low-mass star forming regions (100$\,$pc) and is inclined at 30$^\circ$ to edge on. From these observations integrated intensity, intensity weighted velocity and position velocity diagrams through the centre of the model were created.
The simulations done are focused upon frequencies within ALMA band 7, selected to give the best trade off between resolution and sensitivity for early ALMA science.
(Note the moment 1 and 0 maps were created by integrating between -12.5 to -0.5 km$\,$s$^{-1}$ and +0.5 to +12.5 km$\,$s$^{-1}$ to avoid being dominated by the contribution from the envelope, this can be seen in some PV diagrams as the strong absorption feature at all positions around zero velocity, moment 1 maps are shown with a cutoff of 1/100 of the peak emission/absorption value)\newline

@@

\begin{figure*}
 \includegraphics[width=168mm]{Figures/sim/imageOCS_28-27_30deg_composite_all.eps}
 \label{OCS_all}
 \caption{OCS 28-27 {\bf Left:} Continuum subtracted integrated intensity map, {\bf Right:} Intensity weighted velocity map}
\end{figure*}

\begin{figure}
 \includegraphics[width=84mm]{Figures/sim/imageOCS_28-27_30deg_composite_PV_centre.eps}

 \caption{OCS 28-27 PV through centre}
\end{figure}

After looking at a variety of lines which could be observed of ALMA band 6 or 7 (211-275$\,$GHz and 275-373$\,$GHz) we found that the OCS lines in alma band 7 (22-21 through 30-29) with upper energy levels between 161 and 271 K above the ground state provide a way to trace hot shocked gas in spiral arms without resolving structure. Figures \ref{OCS_mom0} \& \ref{smoothOCS_mom0} show the integrated intensity of the OCS J=28-27 line in the model described previously and one where the disc section of the model is an axis-symmetric model created by taking the azimuthal average of density, temperature and molecular abundance. The order of magnitude difference in the line intensity provides a method to detect hot dense gas without being able to resolve the spiral structure spatially.\newline
%
%\begin{figure}
% \includegraphics[width=84mm]{Figures/sim/imagesmoothedOCS_28-27_30deg_contSub.png}
% \label{smoothOCS_mom0}
% \caption{SmoothedOCS 28-27 Continuum subtracted mom0}
%\end{figure}

@@

\begin{figure}
 \includegraphics[width=84mm]{Figures/sim/imageH2CO_4-0-4--3-0-3_30deg_composite_PV_centre.eps}

 \caption{H$_2$CO 4$_{0\,4}$ - 3$_{\,0\,3}$ PV through centre}
\end{figure}

More complex molecules such as H$_2$CO with many closely spaced spectral lines can be used to gain an estimate on the temperature of a region.\newline

@@

\begin{figure}
 \includegraphics[width=84mm]{Figures/sim/imageHCOp_1-0_30deg_composite_PV_centre.eps}

 \caption{HCO$^+$ 1-0 pv through centre, with the rotation curve from figure \ref{velocity} for comparison}
\end{figure}

@@

In all the simulations except HCO$^+$ we see molecular lines in absorption throughout the majority of the disc and in some cases a small amount of emission towards the centre. Molecular lines throughout the plane of the disc show up in absorption against the continuum emission of the shock heated, dense midplane.

@@

In order to make predictions on the observability of the synthetic brightness maps created by LIME, CASA (common astronomy software applications) was used to simulate observing these with the completed ALMA in it's most extended comfiguration for a total time of four hours.

@@

@@

\section{Grid Construction} \label{sec:gridding} 
In order to construct the grid, candidate points are randomly selected from the volume to be simulated. These candidates then have their density and molecular density compared against a reference point in order to decide if the point is to be used in the grid or not. Candidate grid points are selected at random in cylindrical co-ordinates, linearly spaced in z and $\phi$ and logarithmically spaced in r. For each point to be selected, a random number $\alpha$ is drawn from the semi-open set [0,$\,$1) as a threshold. After selection of random co-ordinates, the hydrogen density and molecular density at the candidate point (n and m, respectively) are compared against the densities of a reference point on the inner edge of the disc (n$_0$ and m$_0$). If $\alpha<\left( \frac{n}{n_0} \right)^{0.3}$ or $\alpha< \left( \frac{m}{m_0} \right)^{0.3}$ then the point is selected for use, otherwise another r, $\phi$, z co-ordinate is selected and this becomes the candidate point. The function comparing the candidate point to the reference point and the candidate point distribution were selected empirically to sample all the scales while ensuring that the majority of points are located in the inner disc where the density is higher. 20\% of these points are forced to be at radii greater than $\sqrt{R_{min}R_{max}}$ (where $R_{min}$ and $R_{max}$ are the inner and outer radius of the model) in order to stop too many of the selecected points clustering in the high density disc and leaving the envelope undersampled. In addiation to this method of selection, 5\% of the points are linearly distributed in x, y and z with no bias with regards to density or abundance. This provides a minimum level of sampling for the large low density regions in the outer parts of the simulated volume. See figure \ref{points} for an example of the points distribution in r, z. \newline

@@

Figures showing different inclinations  
\begin{figure}
 \includegraphics[width=84mm]{Figures/sim/imageC18O_3-2_15deg_contSub.eps}

>>>>

\title[Young protoplanetary discs]{Simulated observations of young protoplanetary discs}
%authors arent splitting over 2 lines so rawlings is pushed off the edge
\author[Tom Douglas, Paola Caselli, Et al.]{Tom Douglas$^{1}$\thanks{E-mail:
pytd@leeds.ac.uk}, Paola Caselli$^{1}$, Aaron Boley$^{2}$, Richard Durisen$^{3}$, Tom Hartquist$^{1}$, John Ilee$^{1}$, Jonathon Rawlings$^{4}$ \\
$^{1}$School of Physics and Astronomy, University of Leeds, Leeds LS2 9JT, UK \\
$^{2}$Department of Astronomy, University of Florida, 211 Bryant Space Center, PO Box 112055, USA\\
$^{3}$Department of Astronomy, Indiana University, 727 East 3rd Street, Swain West 319, Bloomington, IN 47405, USA\\
%$^{4}$Department of Physics & Astronomy, University College London, London WC1E 6BT, UK\\
}
\begin{document}

@@

The formation and early evolution of protoplanetary discs around solar-type and low-mass protostars has little observational support, despite of the fast growing list of theoretical models on the dynamical evolution of star forming dense cores (e.g. Krasnopolsky et al. 2011; Machida et al. 2011; Braiding \& Wardle 2012). The reason for this is that young protostars are surrounded by thick envelopes and power energetic outflows, thus observations of the young discs, predicted to have sizes of about 100\,AU and masses as large as 10\% the original core mass (e.g. Joos et al. 2012), are challenging. The use of sensitive interferometers is needed to achieve high angular resolution and spatially/spectrally disentangle the various disc, envelope and outflow components as well to filter out the extended emission tracing envelope material.

After the pioneer work of, e.g., Chandler et al. (1995), Brown et al. (2000), Looney et al. (2000), recent interferometric observations have discovered compact embedded discs in a sample of Class 0 (Andr\'e et al. 2000) sources, finding masses between 0.4 and  $>$1 M$_{\odot}$ (Jorgensen et al. 2007, 2009; Enoch et al. 2011). A 130\,AU disc was discovered toward a Class 0 source in Perseus, using Jansky Very Large Array (JVLA) observations of NH$_3$ (Choi et al. 2007). Pineda et al. (2012) observed methyl formate with the Atacama Millimetre and millimetre Array (ALMA) and found evidence of rotation toward one of the proto-binary Class 0 sources embedded in IRAS 16293-2422. These observations are consistent with an almost edge-on disc. Persson et al. (2012) observed H$_2^{18}$O with ALMA toward the same source and found evidence of relatively large ($\ga$100\,K) excitation temperatures, as well as a HDO/H$_2$O abundance ratio close to that measured on Earth's oceans and Jupiter-family comets (Hartogh et al. 2011). When ALMA will be completed, it will be finally possible to spatially resolve these young discs and, for the first time, put stringent constraints on theoretical models. As full-operational ALMA is fast approaching, it is important to provide observational predictions based on dynamical models of young protoplanetary discs.

Previous work on simulating observations of gravitationally unstable discs have performed calculations of the continuum radiative transfer in order to look at the structure of the disc to detect disc fragmentation (Ruge et al. 2012) or have used the local thermodynamic equilibrium (LTE) approximation in order to estimate the molecular line emission for an unembedded disc (Krumholtz et al. 2007).

In this paper, we focus our attention to the young self-gravitating discs of Boley \& Durisen (2008), where episodic heating induced by spiral shocks is present, as this may be a good representation of the earliest phases of protoplanetary discs and an alternative to the young ``static'' discs studied by, e.g. Visser et al. (2009, 2011). As shown by Ilee et al. (2011), the spiral shocks cause desorption of volatiles from dust icy mantles and trigger gas-phase chemical reactions with activation energies, too high to occur at lower temperatures. These processes produce clear chemical signatures of the disc dynamics. With the use of 3D radiative transfer modelling, we perform simulated ALMA observations of the disc studied by Ilee et al. (2011) and identify the best tracers of the physical structure of self-gravitating discs. The physical, chemical and radiative transfer models are described in Sect\,\ref{sec:description_model}. Radiative transfer results are in Sect.\,\ref{sec:model_results}, while ALMA simulated observations are in Sec.\,\ref{sec:alma_predictions}. Discussions and conclusions can be found in Sect.\,\ref{sec:discussion}.

@@

The physical model used to simulate the emission from a young protoplanetary disc is a hybrid model obtained from embedding a gravitationally unstable protoplanetary disc  within an envelope with characteristics similar to the well-studied pre-stellar core L1544, as derived by Keto \& Caselli (2010; hereafter KC2010). The KC2010 model follows the dynamical, chemical and thermal evolution of a contracting Bonnor-Ebert sphere (Bonnor 1956; Ebert 1957) with total mass of 10\,M$_{\odot}$, until it reaches the density, temperature and velocity profiles which best match observations. The pre-stellar core model adopted here contains slight modifications due to the inclusion of oxygen cooling in the outer regions of the cloud, where CO is mostly photodissociated (Keto et al., in preparation), and it is shown in Fig.\,\ref{fig:l1544_model}. \newline

\begin{figure}
 \includegraphics[width=84mm]{Figures/model/L1544model_used_legend_small.eps}
 \label{fig:l1544_model}
 \caption{Model of the pre-stellar core L1544 used as the envelope of the young protoplanetary disc in the hybrid model. Showing temperature (red) and dust (green) temperature in kelvin, log number density (blue) in cm$^{-3}$ and inward velocity$\times$100 (cyan) in m$\,$s$^{-1}$. Adapted from Keto \& Caselli (2010) and Keto et al. (2012, in preparation).}
\end{figure}

\begin{figure*}
 \includegraphics[width=168mm]{Figures/model/rhoT2.eps}
 \label{rhoT} 
 \caption{{\bf Left:} A 3D plot of log number density (cm$^{-3}$) showing the spiral structure in the xy plane and scale height of the disc. {\bf Right:} The 3D temperature structure of the disc; regions cooler than 40 degrees are not shown in 3D, demonstrating the narrow central region containing hot material.}
\end{figure*}

@@

%-describe the chemistry  - refer to Ilee et al. which have been taken as input to the rad transfer (RT) code
Chemical abundances in the disc were taken from Ilee et. al (2011) which followed the changes of chemical abundances of trace particles moving through the disc as it evolved. The abundances of 125 species related by 1334 reactions were followed through the time evolution of the disc. These abundances were interpolated onto a 51$^3$ grid covering the disc with cells of size 2.2$\times$2.2$\times$0.22$\,$au.  From Ilee et al. (2011) we selected the three species which appear to trace different regions of the disc : the inner 30\,AU (OCS), the inner 60\,AU (H$_2$CO) and the region between $\simeq$20 and 60\,AU (HCO$^+$) and one which traces the entirety of the disc (C$^{18}$O). As explained by Ilee et al. (2011),  H$_2$CO and OCS mostly probe the central warm regions, where icy mantles evaporate, whereas HCO$^+$ preferentially traces the outer spiral pattern as in the central region it is destroyed by water molecules. The simple chemistry in the KC2010 model, adopted here as the envelope of the protoplanetary disc, does not provide detailed abundances of molecular species (besides CO and H$_2$O, see also Caselli et al. 2012). As discussed in the result section (Sec.\,\ref{sec:model_results}), rough guesses have been made based on values measured toward similar objects.

\subsection{The radiative transfer code} \label{subsec:radiative_transfer_code}
%-describe the RT used (LIME) 
%-describe the RT used (LIME) 
The radiative transfer program used, LIME (Brinch \&Hogerheijde 2010), calculates line intensities based on a weighted sample of randomly chosen points in a continuous 3D model. The method of selecting these points is given in the grid-construction section. At each of these points, the density of the main collision partner (H$_2$), gas and dust temperatures, velocity, molecular abundances and turbulent velocity are taken from the model. These points are then smoothed by Lloyds algorithm (Lloyd 1982) in order to minimise the variation in distance between points whilst keeping the same underlying distribution. These points are then connected by Delaunay triangulation \footnote{In three dimensions this means that if four points are connected into a tetrahedron, the sphere circumscribing these four points contains no other points. It can be shown that this connection is unique for a given set of points.} and it is down these paths that photon propagation is restricted (figures \ref{grid}). The level populations of the selected molecules are calculated at each of these points from collisional and radiative (de)excitation and the local radiation field is calculated. This is repeated 20 times with the populations of each level converging towards a single value. This number of iterations is sufficient for the signal to noise ratio of the level populations (as defined in Brinch \&Hogerheijde 2010) to exceed 1000 in 99\% of the points, ensuring that the simulation has converged on a stable level population. \newline
After 20 iterations of updating the level populations the model is ray-traced in order to produce synthetic brightness maps. In order to minimise the artifacts in the output images, resulting from the grid construction, the average of ten separate runs was taken (figure \ref{averages}).

%\begin{figure}
% \includegraphics[width=84mm]{Figures/sim/average_comparison.eps}
% \label{averages}
% \caption{A plot showing the difference between a single LIME continuum image at 300$\,$GHz (right) and the average of ten such images from different runs of the same model in LIME (left).}
%\end{figure}

@@

\begin{figure}
 \includegraphics[width=84mm]{Figures/model/lime_points_rz_histo2.eps}
 \label{points}
 \caption{A 2D histogram of the point distribution throughout the model. The disc and envelope can be seen as two separate entities which have to be sampled using different point distributions}
\end{figure}

@@

Simulations are limited to molecules for which we have abundances in both the disc and envelope, and also have calculated Einstein and collisional coefficients. The Molecular data used is from the lamda database (reference? http://home.strw.leidenuniv.nl/$\sim$moldata/). Lines for the molecules which satisfy these conditions were simulated with LIME for frequencies between 50$\sim$500$\,$GHz, with a focus on lines which could be observed in ALMA band 7 and molecules which are not typically seen in outflows. The results presented in this section represent those lines which show detectable emission/absorption which can be used to trace either spiral structure or rotation.\newline

Transitions focused upon are CO 3-2 (and isotopologues) at 345.8GHz with an upper energy level of 33.2K, HCO+ 1-0 at 89.2GHZ and 4.3K, OCS 28-27 at 340.4GHz and 237.0K and H2CO $4_{04}$-3$_{03}$ at 290.6GHz and 34.9K {\bf maybe worth including a higher frequency line as well? CS at ~650GHz?}. Other molecules simulated but not shown include HCN, HNC, HNO, HCS$^+$ and CS.\newline

For the purpose of simulating observations the model was placed at roughly the distance of nearby low-mass star forming regions (100$\,$pc) and is inclined at 30$^\circ$ to edge on. From these observations integrated intensity, intensity weighted velocity and position velocity diagrams through the centre of the model were created.
The simulations done are focused upon frequencies within ALMA band 7, selected to give the best trade off between resolution and sensitivity for early ALMA science.
(Note the moment 1 and 0 maps were created by integrating between -12.5 to -0.5 km$\,$s$^{-1}$ and +0.5 to +12.5 km$\,$s$^{-1}$ to avoid being dominated by the contribution from the envelope, this can be seen in some PV diagrams as the strong absorption feature at all positions around zero velocity, moment 1 maps are shown with a cut-off of 1/100 of the peak emission/absorption value)\newline

@@

\begin{figure*}
 \includegraphics[width=168mm]{Figures/sim/imageOCS_28-27_30deg_composite_all.eps}
 \label{OCS_all}
 \caption{OCS 28-27 {\bf Left:} Continuum subtracted integrated intensity map showing clearly differentiated spiral structure in the inner regions of the disc, {\bf Right:} Intensity weighted velocity map}
\end{figure*}

%\begin{figure}
% \includegraphics[width=84mm]{Figures/sim/imageOCS_28-27_30deg_composite_PV_centre.eps}
% \caption{OCS 28-27 position-velocity diagram through centre}
%\end{figure}

OCS traces only the innermost region of the disc and can be used to examine the central regions. As OCS is not seen in outflows it can be used to view rotation of the central part of the disc without risk of contamination. As the excitation energy for the high J transitions of OCS is much higher than the other lines under consideration it is only excited in the hottest densest regions. As a result of this the spiral structure is clearer in OCS than in most transitions, with the low density regions between spiral arms showing no emission or absorption(see figure \ref{ocs_all}). The majority of the OCS is visible in absorption against the bright continuum emission from the disc midplane. The exception to this in the centre of the disc and is an effect of the inclination angle, whereby the line of sight passes through gas above the midplane and then through the hole in the centre of the disc.

%\begin{figure}
% \includegraphics[width=84mm]{Figures/sim/imagesmoothedOCS_28-27_30deg_contSub.png}
% \label{smoothOCS_mom0}
% \caption{SmoothedOCS 28-27 Continuum subtracted mom0}
%\end{figure}

@@

%\begin{figure}
% \includegraphics[width=84mm]{Figures/sim/imageH2CO_4-0-4--3-0-3_30deg_composite_PV_centre.eps}
% \caption{H$_2$CO 4$_{0\,4}$ - 3$_{\,0\,3}$ PV through centre}
%\end{figure}

The abundance H$_2$CO molecule is fairly typical of the molecules simulated in Ilee (2011) in that it traces the spiral structure roughly the inner 2/3$^{rd}s$ of the disc. The spiral structure of the disc can be seen clearly in the integrated intensity map and its rotation of detected out to larger radii than is possible with the OCS and is more visible between spiral arms but is otherwise simimlar in morphology.

@@

%\begin{figure}
% \includegraphics[width=84mm]{Figures/sim/imageHCOp_1-0_30deg_composite_PV_centre.eps}
% \caption{HCO$^+$ 1-0 pv through centre, with the rotation curve from figure \ref{velocity} for comparison}
%\end{figure}

@@

In all the simulations except HCO$^+$ we see molecular lines in absorption throughout the majority of the disc and in some cases a small amount of emission towards the centre. Molecular lines throughout the plane of the disc show up in absorption against the continuum emission of the shock heated, dense mid-plane.

@@

In order to make predictions on the observability of the synthetic brightness maps created by LIME, CASA (common astronomy software applications) was used to simulate observing these with the completed ALMA in its most extended comfiguration for a total time of four hours.

@@

PV diagrams allow us to reconstruct the rotation of the disc. ALMA has enough resolution and sensitivity to see spiral structure clearly in continuum emission but also in CO isotopologues and marginally in OCS.

@@

\section{Grid Construction} \label{sec:gridding}

In order to construct the grid, candidate points are randomly selected from the volume to be simulated. These candidates then have their density and molecular density compared against a reference point in order to decide if the point is to be used in the grid or not. Candidate grid points are selected at random in cylindrical co-ordinates, linearly spaced in z and $\phi$ and logarithmically spaced in r. For each point to be selected, a random number $\alpha$ is drawn from the semi-open set [0,$\,$1) as a threshold. After selection of random co-ordinates, the hydrogen density and molecular density at the candidate point (n and m, respectively) are compared against the densities of a reference point on the inner edge of the disc (n$_0$ and m$_0$). If $\alpha<\left( \frac{n}{n_0} \right)^{0.3}$ or $\alpha< \left( \frac{m}{m_0} \right)^{0.3}$ then the point is selected for use, otherwise another r, $\phi$, z co-ordinate is selected and this becomes the candidate point. The function comparing the candidate point to the reference point and the candidate point distribution were selected empirically to sample all the scales while ensuring that the majority of points are located in the inner disc where the density is higher. 20\% of these points are forced to be at radii greater than $\sqrt{R_{min}R_{max}}$ (where $R_{min}$ and $R_{max}$ are the inner and outer radius of the model) in order to stop too many of the selected points clustering in the high density disc and leaving the envelope under sampled. In addition to this method of selection, 5\% of the points are linearly distributed in x, y and z with no bias with regards to density or abundance. This provides a minimum level of sampling for the large low density regions in the outer parts of the simulated volume. See figure \ref{points} for an example of the points distribution in r, z. \newline

@@

Figures showing different inclinations {\bf (will redo these for compound images soon)}
\begin{figure}
 \includegraphics[width=84mm]{Figures/sim/imageC18O_3-2_15deg_contSub.eps}
